---
title: "Plots for hw02"
author: "Cole Brookson"
date: "`r Sys.Date()`"
output: html_document
---

# Question 1b

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```{r}
# Given data
y_obs <- c(1.7, 5.3, 2.1)  # Observations
b <- 0.5  # Known scale parameter
n <- length(y_obs)  # Number of observations

# Prior parameters (weakly informative)
alpha_0 <- 1
beta_0 <- sum(exp(b * y_obs)) / (2 * n)

# Posterior parameters
alpha_post <- alpha_0 + n
beta_post <- beta_0 - sum(1 - exp(b * y_obs))

# Posterior distribution (Gamma distribution)
posterior_mean <- alpha_post / beta_post
posterior_median <- qgamma(0.5, shape = alpha_post, rate = beta_post)
posterior_sd <- sqrt(alpha_post) / beta_post
posterior_quantiles <- qgamma(c(0.025, 0.5, 0.975), shape = alpha_post, rate = beta_post)

# Display key summary statistics
posterior_summary <- list(
  posterior_mean = posterior_mean,
  posterior_median = posterior_median,
  posterior_sd = posterior_sd,
  posterior_quantiles = posterior_quantiles
)
posterior_summary

```

```{r}
# Generate a histogram for the posterior distribution
x_vals <- seq(qgamma(0.001, shape = alpha_post, rate = beta_post), 
              qgamma(0.999, shape = alpha_post, rate = beta_post), 
              length.out = 1000)

posterior_pdf <- dgamma(x_vals, shape = alpha_post, rate = beta_post)
hist(posterior_pdf)
```

# Question 2c

```{r}
# Observation
Y <- 4.3

# Compute posterior probabilities
posterior_probs <- numeric(10)
for (j in 1:10) {
  posterior_probs[j] <- 1 / (1 + (Y - j)^2)
}
posterior_probs <- posterior_probs / sum(posterior_probs)  # Normalize

# Compute posterior mean
posterior_mean <- sum(1:10 * posterior_probs)
print(paste0("The posterior mean is ", posterior_mean))
```